Online Multiclass Linear Classification with Bandit Feedback

### 2018-11-21 arxiv-version/main.tex
Finish lower bound proof

We design new algorithms for the problem of \textsc{Online Multiclass
Classification with Bandit Feedback}. In this problem, in each round, the
algorithm receives a feature vector and algorithm has to predict one of $K$
possible classes. Imediately after the algorithm makes its prediction, it
receives feedback whether the prediction was correct or not. However, in
contrast with the standard, so called full information setting, the algorithm
does \emph{not} receive the correct label. The feedback is only binary. To allow
for meaningful randomized strategies, we assume that the correct label in a
particular round is chosen prior to algorithm's prediction in that
round.\footnote{The reader can assume throughout the paper a simpler model of
the so called \emph{oblivious adversary} where the whole sequence of labeled
examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is chosen before the first
round of the game.} The protocol of the problem is stated below.

@@

For the case when the examples are strongly linearly separable, we design a
simple and efficient algorithm with expected number of mistakes at most $(K-1)
\lfloor (R/\gamma)^2 \rfloor$. The algorithm can be viewed as running $K$ copies
of the \textsc{Binary Perceptron} algorithm, one copy for each class.
Intuitively speaking, the factor $K-1$ is the price we pay for the bandit
feedback, or more precisely, the lack of full-information feedback. We also
prove that any algorithm makes $K \lfloor (R/\gamma)^2 \rfloor / ??$ in the
worst case in this setting.

@@

We define two notions of linear separability for multiclass classification. The
first notion is the standard notion of linear separability used in the proof of
mistake bound for the \textsc{Multiclass Perceptron} algorithm. The second
notion is stronger, i.e. more restrictive. However, it is more suitable for the
bandit setting, since it allows for a simple and efficient algorithm; see
Section~\ref{section:algorithm-for-strongly-linearly-separable-data}.

@@

\begin{algorithm}[h]
\caption{\textsc{Bandit Algorithm for Strongly Linearly Separable Examples}
\label{algorithm:algorithm-for-strongly-linearly-separable-examples}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$. Inner product space $(V,\ip{\cdot}{\cdot})$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t \in V$}
\STATE{Compute $S_t = \left\{ i ~:~ 1 \le i \le K, \ \ip{w_i^{(t)}}{x_t} \ge 0 \right\}$}
\IF{$S_t = \emptyset$}
\STATE{Predict $\widehat y_t \sim \text{Uniform}(\{1,2,\dots,K\})$}
\ELSE
\STATE{Predict $\widehat y_t \sim \text{Uniform}(S_t)$}
\ENDIF
\STATE{Observe feedback $z_t \in \{0,1\}$ where $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$S_t = \emptyset$}
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} + x_t$}
\ENDIF
\ELSE
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

@@

\begin{theorem}[Mistake lower bound]
\label{theorem:strongly-separable-example-mistake-lower-bound}
Let $\gamma$ be a positive real number and let $R$ be a non-negative real
number. For any algorithm for the \textsc{Online Multiclass Classification with
Bandit Feedback} with $K \le \frac{1}{2}(R/\gamma)^2$ classes there exists an inner
product space $(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a
sequence of labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$
in $V \times \{1,2,\dots,K\}$ that are strongly linearly separable with
margin $\gamma$, the norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T}
\le R$ and the expected number of mistakes the algorithm makes is at least $K
\lfloor \frac{1}{8} (R/\gamma)^2 \rfloor/ ??$.
\end{theorem}

\begin{proof}
Let $M = \lfloor \frac{1}{8} (R/\gamma)^2 \rfloor$. Let $V = \R^{M+1}$ equipped with the
standard inner product.  Let $e_1, e_2, \dots, e_{M+1}$ be the standard
orthonormal basis of $\R^{M+1}$. We define vectors $v_1, v_2, \dots, v_M \in V$
where $v_i = \frac{R}{\sqrt{2}}(e_i + e_{M+1})$ for $i=1,2,\dots,M$. Let $\ell_1, \ell_2,
\dots, \ell_M$ be chosen i.i.d. uniformly at random from $\{1,2,\dots,K\}$ and
independently of any randomness used the by the algorithm.

Let $T = M K$. We define examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ as
follows:
$$
\begin{array}{lclclclcl}
(x_1, y_1) & = & (x_2, y_2) & = & \cdots & = & (x_K, y_K) & = & (v_1, \ell_1) \; , \\
(x_{K+1},y_{K+1}) & = & (x_{K+2}, y_{K+2}) & = & \cdots & = & (x_{2K}, y_{2K}) & = & (v_2, \ell_2) \; , \\
& \vdots & \\
(x_{(M-1)K+1}, y_{(M-1)K+1})  & = & (x_{(M-1)K+2}, y_{(M-1)K+2}) & = & \cdots & = & (x_{MK}, y_{MK}) & = & (v_M, \ell_M) \; .
\end{array}
$$

Clearly, norm of each example is $R$. We show that the examples are strongly
separable with margin $\gamma$. In order to see that consider $w_1^*, w_2^*,
\dots, w_K^* \in V$ defined by
$$
w_i^* = 2 \frac{\gamma}{R} \left( \sum_{j ~:~ \ell_j = i} e_j \right) - \frac{\gamma}{R} e_{M+1} \qquad \quad \text{for $i=1,2,\dots,K$.}
$$
Since for any $i \in \{1,2,\dots,K\}$ and any $j \in \{1,2,\dots,M\}$,
$$
\ip{w_i^*}{v_j} =
\begin{cases}
\gamma & \text{if $\ell_j = i$,} \\
- \gamma & \text{if $\ell_j \neq i$,}
\end{cases}
$$
it means that $w_1^*, w_2^*, \dots, w_K^*$ satisfy
\eqref{equation:strong-linear-separability-2} and
\eqref{equation:strong-linear-separability-3}. Condition \eqref{equation:strong-linear-separability-1}
is satisfied since
$$
\sum_{i=1}^K \norm{w_i^*}^2
= 4 \frac{\gamma^2}{R^2} \sum_{j=1}^M \norm{e_j}^2 + \frac{\gamma^2}{R^2} K \norm{e_{M+1}}^2
= 4\frac{\gamma^2}{R^2} M + \frac{\gamma^2}{R^2} K
\le \frac{1}{2} + \frac{1}{2}
= 1 \; .
$$

It remains to show that the expected number of mistakes the algorithm makes is
at least $K M / ??$.

TODO
\end{proof}

>>>>

We design new algorithms for the problem of \textsc{Online Multiclass
Classification with Bandit Feedback}. In this problem, in each round, the
algorithm receives a feature vector and the algorithm has to predict one of $K$
possible classes. Imediately after the algorithm makes its prediction, it
receives feedback whether the prediction was correct or not. However, in
contrast with the standard, so called full information setting, the algorithm
does \emph{not} receive the correct label. The feedback is only binary. To allow
for meaningful randomized strategies, we assume that the correct label in a
particular round is chosen prior to algorithm's prediction in that
round.\footnote{The reader can assume throughout the paper a simpler model of
the so called \emph{oblivious adversary} where the whole sequence of labeled
examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$ is chosen before the first
round of the game.} The protocol of the problem is stated below.

@@

For the case when the examples are strongly linearly separable, we design a
simple and efficient algorithm with expected number of mistakes at most $(K-1)
\lfloor (R/\gamma)^2 \rfloor$. The algorithm can be viewed as running $K$ copies
of the \textsc{Binary Perceptron} algorithm, one copy for each class.
Intuitively speaking, the factor $K-1$ is the price we pay for the bandit
feedback, or more precisely, the lack of full-information feedback. We also
prove that any (possibly randomized) algorithm makes $\Omega(K (R/\gamma)^2))$
mistakes in the worst case.

@@

We define two notions of linear separability for multiclass classification. The
first notion is the standard notion of linear separability used in the proof of
the mistake bound for the \textsc{Multiclass Perceptron} algorithm. The second
notion is stronger, i.e. more restrictive. However, it is more suitable for the
bandit setting, since it allows for a simple and efficient algorithm; see
Section~\ref{section:algorithm-for-strongly-linearly-separable-data}.

@@

\begin{algorithm}[h]
\caption{\textsc{Bandit Algorithm for Strongly Linearly Separable Examples}
\label{algorithm:algorithm-for-strongly-linearly-separable-examples}}
\begin{algorithmic}[1]
{
\REQUIRE{Number of classes $K$. Number of rounds $T$. Inner product space $(V,\ip{\cdot}{\cdot})$.}
\STATE{Initialize $w_1^{(1)} = w_2^{(1)} = \dots = w_K^{(1)} = 0$}
\FOR{$t=1,2,\dots,T$}
\STATE{Observe feature vector $x_t \in V$}
\STATE{Compute $S_t = \left\{ i ~:~ 1 \le i \le K, \ \ip{w_i^{(t)}}{x_t} \ge 0 \right\}$}
\IF{$S_t = \emptyset$}
\STATE{Predict $\widehat y_t \sim \text{Uniform}(\{1,2,\dots,K\})$}
\ELSE
\STATE{Predict $\widehat y_t \in S_t$ chosen arbitrarily}
\ENDIF
\STATE{Observe feedback $z_t \in \{0,1\}$ where $z_t = \indicator{\widehat y_t \neq y_t}$}
\IF{$S_t = \emptyset$}
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} + x_t$}
\ENDIF
\ELSE
\IF{$z_t = 1$}
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\} \setminus \{\widehat y_t\}$}
\STATE{Update $w_{\widehat y_t}^{(t+1)} = w_{\widehat y_t}^{(t)} - x_t$}
\ELSE
\STATE{Set $w_i^{(t+1)} = w_i^{(t)}$ for all $i \in \{1,2,\dots,K\}$}
\ENDIF
\ENDIF
\ENDFOR
}
\end{algorithmic}
\end{algorithm}

@@

\begin{theorem}[Mistake lower bound]
\label{theorem:strongly-separable-example-mistake-lower-bound}
Let $\gamma$ be a positive real number and let $R$ be a non-negative real
number. For any algorithm for the \textsc{Online Multiclass Classification with
Bandit Feedback} with $K \le \frac{1}{4}(R/\gamma)^2$ classes there exists an inner
product space $(V, \ip{\cdot}{\cdot})$, a non-negative integer $T$ and a
sequence of labeled examples $(x_1, y_1), (x_2, y_2), \dots, (x_T, y_T)$
in $V \times \{1,2,\dots,K\}$ that are strongly linearly separable with
margin $\gamma$, the norms satisfy $\norm{x_1}, \norm{x_2}, \dots, \norm{x_T}
\le R$ and the expected number of mistakes the algorithm makes is at least $\frac{K-1}{2}
\lfloor \frac{1}{16} (R/\gamma)^2 \rfloor$.
\end{theorem}

\begin{proof}
We use probabilistic method. Let $M = \lfloor \frac{1}{16} (R/\gamma)^2 \rfloor$.
Let $V = \R^{M+1}$ equipped with the standard inner product.  Let $e_1, e_2,
\dots, e_{M+1}$ be the standard orthonormal basis of $V$. We define
vectors $v_1, v_2, \dots, v_M \in V$ where $v_i = \frac{R}{\sqrt{2}}(e_i +
e_{M+1})$ for $i=1,2,\dots,M$. Let $\ell_1, \ell_2, \dots, \ell_M$ be chosen
i.i.d. uniformly at random from $\{1,2,\dots,K\}$ and independently of any
randomness used the by the algorithm. Let $T = M K$. We define examples $(x_1,
y_1), (x_2, y_2), \dots, (x_T, y_T)$ as follows:
$$
\begin{array}{lclclclcl}
(x_1, y_1) & = & (x_2, y_2) & = & \cdots & = & (x_K, y_K) & = & (v_1, \ell_1) \; , \\
(x_{K+1},y_{K+1}) & = & (x_{K+2}, y_{K+2}) & = & \cdots & = & (x_{2K}, y_{2K}) & = & (v_2, \ell_2) \; , \\
& \vdots & \\
(x_{(M-1)K+1}, y_{(M-1)K+1})  & = & (x_{(M-1)K+2}, y_{(M-1)K+2}) & = & \cdots & = & (x_{MK}, y_{MK}) & = & (v_M, \ell_M) \; .
\end{array}
$$

With probability one, norm of each example is $R$. We show that with probability
one, the examples are strongly separable with margin $\gamma$. In order to see
that consider $w_1^*, w_2^*, \dots, w_K^* \in V$ defined by
$$
w_i^* = 2 \sqrt{2} \frac{\gamma}{R} \left( \sum_{j ~:~ \ell_j = i} e_j \right) - \sqrt{2} \frac{\gamma}{R} e_{M+1} \qquad \quad \text{for $i=1,2,\dots,K$.}
$$
Since for any $i \in \{1,2,\dots,K\}$ and any $j \in \{1,2,\dots,M\}$,
$$
\ip{w_i^*}{v_j} =
\begin{cases}
\gamma & \text{if $\ell_j = i$,} \\
- \gamma & \text{if $\ell_j \neq i$,}
\end{cases}
$$
it means that $w_1^*, w_2^*, \dots, w_K^*$ satisfy
\eqref{equation:strong-linear-separability-2} and
\eqref{equation:strong-linear-separability-3}. Condition \eqref{equation:strong-linear-separability-1}
is satisfied since
$$
\sum_{i=1}^K \norm{w_i^*}^2
= 8 \frac{\gamma^2}{R^2} \sum_{j=1}^M \norm{e_j}^2 + 2 \frac{\gamma^2}{R^2} K \norm{e_{M+1}}^2
= 8 \frac{\gamma^2}{R^2} M + 2 \frac{\gamma^2}{R^2} K
\le \frac{1}{2} + \frac{1}{2}
= 1 \; .
$$

It remains to lower bound the expected number of mistakes of the algorithm. For
any $j \in \{1,2,\dots,M\}$, consider the expected number of mistakes the
algorithm makes in rounds $K(j-1) + 1, K(j-1) + 2, \dots, Kj$. In these rounds,
the algorithm is guessing the label $\ell_j$. Since $\ell_j$ is chosen uniformly
at random from the set $\{1,2,\dots,K\}$ and feedback is only binary, the
expected number of mistakes the algorithm makes in these rounds is at least
$\frac{K-1}{2}$. Altogether, the algorithm makes at least $\frac{K-1}{2} M$
mistakes in expectation.

Finally, since strong separability and norm condition hold with probability one,
there exists a particular (i.e. deterministic) sequence of examples for which
the algorithm makes at least $\frac{K-1}{2} M$ mistakes.
\end{proof}
